---
title: "Tono DB preliminary analyses"
author: "Steven Moran and Lilja Maria Sæbø"
date: "2023-06-09"
output: github_document
---

```{r, message=FALSE}
library(tidyverse)
library(knitr)
```


Load the [CLDF data](https://github.com/cldf-datasets/tonodb/).

```{r, message=FALSE}
values <- 
  read_csv(url('https://raw.githubusercontent.com/cldf-datasets/tonodb/main/cldf/values.csv'))
languages <- 
  read_csv(url('https://raw.githubusercontent.com/cldf-datasets/tonodb/main/cldf/languages.csv'))
contributions <- 
  read_csv(url('https://raw.githubusercontent.com/cldf-datasets/tonodb/main/cldf/contributions.csv'))
parameters <- 
  read_csv(url('https://raw.githubusercontent.com/cldf-datasets/tonodb/main/cldf/parameters.csv'))
```

We have this many languages in our sample.

```{r}
nrow(languages)
```

And this many observations.

```{r}
nrow(values)
```

Let's map our data points.

```{r}
ggplot(data=languages, aes(x=Longitude, y=Latitude)) + 
  borders("world", colour="gray50", fill="gray50") + 
  geom_point() +
  theme_bw()
```

Note that 23 data points are removed due to NAs in the data. TODO: fix these by adjusting the Glottocodes in the spreadsheet.

```{r}
languages %>% filter(is.na(Latitude)) %>% kable()
```

Here we can add some color by language family.

```{r}
ggplot(data=languages, aes(x=Longitude, y=Latitude, color=family_id)) + 
  borders("world", colour="gray50", fill="gray50") + 
  geom_point() +
  theme_bw()
```

How many data points per macroarea? (Note again several NAs.)

```{r}
table(languages$Macroarea, exclude=FALSE)
```

Recreate some of the tables.

First merge the tonodb tables.

```{r}
tonodb <- left_join(values, languages, by=c("Language_ID"="ID"))
```

```{r}
x <- tonodb %>% select(Macroarea, Language_ID) %>% distinct() %>% group_by(Macroarea) %>% summarise(Languages = n())
y <- tonodb %>% select(Macroarea, family_id) %>% distinct() %>% group_by(Macroarea) %>% summarize(Families = n())
z <- tonodb %>% select(Macroarea, TriggeringContext) %>% group_by(Macroarea) %>% summarize(`Cases of tonogenesis` = n())

tmp <- left_join(x, y)
tmp <- left_join(tmp, z)

tmp %>% kable()
```

Number of languages in different families.

```{r}
tonodb %>% select(family_id, Language_ID) %>% distinct() %>% group_by(family_id) %>% summarize(Languages = n())

tonodb %>% select(family_id, Language_ID) %>% distinct() %>% arrange(family_id, Language_ID) %>% group_by(family_id) %>% summarize(Languages = str_c(Language_ID, collapse=", ")) %>% kable()

```


TODO:

- Triggering context | Cases of tonogenesis | Number of languages


TODO: How to recreate tables 4 and 5 from Lilja's poster?



